---
title: "Pointers/shortcuts in R with the 'pointr' package"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# knitr::opts_knit$set(root.dir = "C:\\Users\\Joachim\\Desktop\\Dokumente\\500  Projekte\\510  R-Packages\\516  xml2relational\\")
```

## Overview

R's built-in copy-on-modify behavior prevents the user from having two symbols always pointing to the same object. Because pointers, as they are common in other programming languages, are essentially symbols (variable) related to an object that has already another symbol attached to it, it is clear that pointers do not fit naturally into R's language concept.

However, pointers would indredibly useful, e.g. when you work with complex subsets of dataframes. These complex filtering conditions make the code harder to read and to maintain. For this reason, it would be good to have a kind of 'abbreviation' or 'shortcut' that lets you write such filtering conditions more efficiently.

The **`pointr`** package provides functionality to create pointers to any R object easily, including pointers to subsets/selections from dataframes.

```{r, include=FALSE}
createAccessFun <- function(symbol1, symbol2) {
  varname <- paste0(".zstor", sample(10000000:99999999, 1)) 
  return(paste0("
  .", symbol1, "<- function(arg = NULL) {
    # ", symbol2, "
    if(!is.null(arg)) {
      assign(\"", varname, "\", eval(arg), env = parent.env(environment()))
      eval(parse(text=\"", symbol2, " <- ", varname, "\"), envir = parent.env(environment()))
      rm(",varname,", envir = parent.env(environment()))
    } 
    else {
      return(eval(parse(text=\"", symbol2,"\"), envir = parent.env(environment())))
    }
  }
 "))
}


ptr <- function(symbol1, symbol2) {
  funcode <- createAccessFun(symbol1, symbol2)
  eval(parse(text = funcode), envir = parent.frame())
  makeActiveBinding(symbol1, get(paste0(".", symbol1), envir = parent.frame()), env = parent.frame())
}


rm.ptr <- function(symbol1, keep = FALSE) {
  rm(list=paste0(".", symbol1), envir=parent.env(environment()))
  if(keep == FALSE) {
    rm(list=symbol1, envir=parent.env(environment()))
  } else {
    pres <- eval(parse(text=symbol1), envir=parent.env(environment()))
    rm(list=paste0(symbol1), envir=parent.env(environment()))
    assign(symbol1, pres, envir=parent.env(environment()))
  }
}


where.ptr <- function(symbol1) {
  code <- capture.output(eval(parse(text=paste0(".", symbol1))))
  return(substring(code[2], stringr::str_locate(code[2], "# ")[2]+1, nchar(code[2])))
}
```



## Working with `pointr`


### Installing and loading `pointr`

To install the CRAN version of **`pointr`** from the R console, just call:

``` {r eval=FALSE}
install.packages("pointr", dependencies = TRUE)
```

Before using **`pointr`**, it needs to be attached to the package search path:
``` {r eval=FALSE}
library(pointr)
```

Now, we are ready to go.


### Functions

From the user's perspective, **`pointr`** provides three simple functions:

* **`ptr(symbol1, symbol2)`** creates a pointer called `symbol1` to the object in `symbol2`. The function has no return value. The `symbol1` pointer variable is created by the function. Both arguments, `symbol1` and `symbol2`, are strings.

* **`rm.ptr(symbol1, keep=TRUE)`** removes the pointer. It deletes the hidden access function `.symbol1()`. If `keep == FALSE` it also deletes the pointer variable `symbol1`. If, however `keep == FALSE` a copy of the object that the pointer refers to is stored in the `symbol1` variable. The `symbol1` argument is a string.

* **`where(symbol1)`** shows the name of the object the pointer `symbol1` points to. The `symbol1` argument is a character vector.

Pointers work like the referenced variable itself. You can, for example, print them (which prints the contents of the referenced variable) or assign values to them (which assigns the values to the referenced variable).


### Examples

#### Example 1: A simple vector

First, we define a variable `myvar` and create a pointer `mypointer` to this variable. Accessing the pointer `mypointer` actually reads `myvar`.
``` {r}
myvar <- 3
ptr("mypointer", "myvar")
mypointer
```
Accordingly, changes to `myvar` can be seen using the pointer.
``` {r}
myvar <- 5
mypointer
```
The pointer can also be used in assignments; this changes the variables the pointer points to:
``` {r}
mypointer <- 7
myvar
```

#### Example 2: Subsetting a dataframe

We create a simple dataframe:
``` {r}
df <- data.frame(list(var1 = c(1,2,3), var2 = c("a", "b", "c")), stringsAsFactors = FALSE)
df
```

Now we set a pointer `sel` to a subset of `df`:
``` {r}
i <- 2
ptr("sel", "df$var2[i]")
```
We can now change...
``` {r}
sel <- "hello"
df$var2[i]
```
and read data from `df` using the `sel` pointer:
``` {r}
df$var2[i] <- "world"
sel
```
We can also check easily where our pointer points to:
``` {r eval=FALSE}
where.ptr("sel")
## df$var2[i]
```
When the index variable `i` changes, our pointer adjusts accordingly:
``` {r}
i <- 3
sel
```



## Technical note

Active bindings are used to create the **`pointr`** pointers. For each pointer an object with active binding is created. Every time the pointer is accessed, the active binding calls a hidden function called `.pointer` where *`pointer`* is the name of the pointer variable. This function evaluates the assignment (if the user assigns a value to the pointer) or evaluates the object the pointer refers to as such (if the user accesses the contents of the object the pointer points to). This way it is possible not only to address objects like vectors or dataframes but also to have pointers to things like, for example, subsets of datafames.

All **`pointr`** functions operate in the environment in which the pointer is created.
